---
title: "Subnational Probabilistic Population Projections"
author: Hana Sevcikova
date: 2023-10-19
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{population-projections}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
library(bayesPop)
```

In this vignette we will show how **bayesPop** can be used in subnational settings, using annual 1x1 data. We will use example data for the US states to demonstrate the various functionalities. For instructions how to use **bayesPop** to generate national projections, see [Ševčíková and Raftery (2016)](https://www.jstatsoft.org/article/view/v075i05).

# Downloading example data

```{r eval = FALSE, include = FALSE}
data_dir <- tempdir()
if(!dir.exists(data_dir))
  dir.create(data_dir)
repo_file <- file.path(data_dir, "main.zip")
download.file("https://github.com/PPgp/bayesPopUSdata/archive/refs/heads/main.zip", repo_file)
unzip(repo_file, data_dir)
```

```{r eval = TRUE, include = TRUE}
data_dir <- "~/bayespop/R/Pop/subnat/data/USA"
```

# Setup



# References

Ševčíková, H. and Raftery, A.E. (2016). [bayesPop: Probabilistic Population Projections](https://www.jstatsoft.org/article/view/v075i05). Journal of Statistical Software, Vol. 75(5).